---
title: "TidyTuesday - Himalayan Mountaineering"
---

The following data set was gathered and provided by TidyTuesday, linked here:\
<https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-01-21/readme.md>

The data was sourced from the Himalayan Database which keeps information on peaks and mountaineering expeditions that take place in the Himalayan Region. The full himalayan database catalog can be found here:

<https://www.himalayandatabase.com/index.html>

The following graph attempts to compare the volume of expeditions per region listed in the data on all mountains exceeding 7,000 meters. This is done by joining the "peaks" and "expeditions" data sets, filtering for only peaks above 7,000m, and then counting the expeditions by region.

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
```

```{r}
tuesdata <- tidytuesdayR::tt_load('2025-01-21')
exped_tidy <- tuesdata$exped_tidy
peaks_tidy <- tuesdata$peaks_tidy
```

```{r, echo=TRUE}
# joining
exped_tidy <- exped_tidy |> mutate(PEAKID = as.character(PEAKID))
peaks_tidy <- peaks_tidy |> mutate(PEAKID = as.character(PEAKID))

# count expeditions for peaks >= 7000m 
exped_by_region <- exped_tidy |>
  distinct(EXPID, PEAKID) |>  
  inner_join(
    peaks_tidy |> select(PEAKID, HEIGHTM, REGION_FACTOR),
    by = "PEAKID"
  ) |>
  filter(HEIGHTM >= 7000) |>
  count(REGION_FACTOR, name = "n_expeditions") |>
  arrange(desc(n_expeditions))

exped_by_region
```

```{r, echo=TRUE}
# ---- plot ----
ggplot(exped_by_region,
       aes(x = forcats::fct_reorder(REGION_FACTOR, n_expeditions),
           y = n_expeditions)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Number of Expeditions on ≥7000 m Peaks, by Region",
    subtitle = "Himalayan Database (TidyTuesday 2025-01-21)",
    x = "Region",
    y = "Number of expeditions"
  ) +
  theme_minimal(base_size = 12)
```
